---
title: "Using powerFlux"
author: "Peter Levy"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using powerFlux}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, eval=FALSE, include = FALSE}
library(rmarkdown)
system.time(render("use_powerFlux.Rmd", output_file = "use_powerFlux.html"))
```

```{r, attachPackages, eval = FALSE, echo = FALSE, warning=FALSE, message=FALSE}
library(here)
library(units)
library(ggplot2)
library(ggforce)
library(viridis)
library(powerFlux)
pkgload::load_all(".")
do_all <- FALSE # TRUE
```

## Introduction 
This document describes the `powerFlux` R package for statistical power analysis and uncertainty quantification for chamber measurements of N$_2$O and CH$_4$ fluxes.  Most users will only use one or two functions `get_ci_flux` and `get_ci_omega`, which return the expected confidence intervals in instantaneous flux measurements and cumulative fluxes expressed as emission factors.

## Installation
The package can be installed directly from a bundled .tar.gz file.
Alternatively, and to keep versions up to date more easily, the package can be installed directly from GitHub, using `install_github` from remotes.

```{r, eval = FALSE}
# If not already installed:
# install.packages(c("remotes"))
library(remotes)  # for install_github

# install powerFlux from github
install_github("NERC-CEH/powerFlux")
```

## Using the package
Now we can load the package.
```{r}
library(powerFlux)
```

and attach some packages it depends upon:
```{r}
library(ggplot2)
library(units)
```

You can then call the `powerFlux` functions directly from the R command line, as described below. The basic input is a set of parameters which define the chamber set-up and sampling characteristics.

## Individual chamber flux measurements
The flux of a GHG from the soil surface within a chamber is calculated from the rate of change in the mixing ratio $d \chi / d t$ (in mol GHG/mol air/ s), adjusted for the density of air per unit surface area:

$$F = \frac{d \chi} {d t} \times \rho \frac{V}{A}$$

where $F$ is the surface flux in mol GHG/m^2^/s, $\rho$ is the density of air in mol/m^3^, and $V$ and $A$ are the volume and area of the chamber (which simplifies to the height $h = V/A$ for most common chamber shapes). The term $\rho h$ can be considered a constant for a given chamber under typical conditions.

The term $d \chi / d t$ is usually estimated as the slope $\beta$ of a linear regression between $\chi$ and time $t$ during the chamber closure (although various methods accounting for non-linearity are also used).
The uncertainty in this term is estimated by the standard algebra of linear regression, where the 95% confidence interval (CI) in the slope $\beta$ is given:

$$CI_{\beta}^{95} = \sqrt{ \frac{\sigma_\chi^2} {n-1 \sum (t_i - \bar{t})^2} } \times \mathbb{T}$$

where $\sigma\chi^2$ is the residual variance, $n$ is the number of data points, $\sum (t_i - \bar{t})^2$ is the variance in the $x$ independent variable, time, and $\mathbb{T}$ represents the t statistic, with a value of 1.96 for a sample size greater than 20. We can define a function which calculates this 95% confidence interval in chamber flux measurements, given the characteristic of the chamber and analyser.

The uncertainty, as quantified by the 95 % confidence interval, depends on the residual (unexplained) variation in mixing ratio $\sigma_\chi$ (`sigma_chi` in the code), the number of gas samples, their spread over time, and the chamber height. We use the `units` package to take care of unit conversions. To estimate $\sigma_\chi$ a priori, we can use the noise or precision quoted for the gas analyser - the standard deviation $\sigma$ in recorded $\chi$ under constant conditions. This will be an underestimate, and there will be additional variation coming from all the vagaries of sampling from a chamber (imperfect mixing, pressure fluctuations, leaks, etc.). Empirically, $\sigma_\chi$ is the residual standard error in the linear regression of $\chi$ versus $t$, as this includes both sources of error, so previous values of this can be used to estimate for future experiments. Chamber height comes into the equation because it determines the scaling factor between changes in mixing ratio and the calculated flux: the same flux will have a smaller effect on the mixing ratio in a tall chamber compared to a short one.

Below, we specify $\sigma_\chi$ of 20 nmol N$_2$O/mol, 
10 gas samples, spread regularly over time 5 minutes, and a chamber height of 23 cm.

```{r, echo = TRUE}
# define the conversion unit between g N and moles of N2O
units::install_unit("mol_n2o", "28 g", "mol wt of N in N2O")

sigma_chi <- set_units(20, nmol_n2o/mol)
height <- set_units(0.23, m)
t_max <- set_units(5 * 60, s)
n_gas <- 10
v_t <- get_v_t(t_max, n_gas)
sigma_t <- get_sigma_t(v_t)

ci <- get_ci_flux(sigma_chi = sigma_chi,
            height = height,
            sigma_t = sigma_t,
            n_gas = n_gas)
print(ci)
```

The 95 % confidence interval for a flux measured with this system will be $\pm$ `r round(ci, 3)` in its associated units.


To illustrate, we can calculate the CI in flux measurements from a chamber of height 0.23 m and an analyser with sigma_chi of 20 nmol/mol (standard deviation) and 10 samples recorded over 5 minutes. 
 
```{r}
meanlog <- 0
sdlog <- 1.5
max_flux <- set_units(5, nmol_n2o / m^2 /s)
get_dt_ci(height = height, sigma_chi = sigma_chi, sigma_t = sigma_t, n_gas = n_gas,
            meanlog = 0, sdlog = 1, max_flux = max_flux)
```

The CI is a constant value given these characteristics. However, we might also consider this in relative terms compared to the flux itself, rather like a signal-to-sigma_chi ratio (SNR).
Therefore, we express this in relation to a range of typical or possible fluxes, up to a maximum value `max_flux`, with their probability of occuring specified by a lognormal distribution, with the parameter `meanlog` and `sdlog`.

```{r}
dt <- get_dt_ci(height = height, sigma_chi = sigma_chi, sigma_t = sigma_t, n_gas = n_gas,
            meanlog = 0, sdlog = 1, max_flux = max_flux)
p <- ggplot(dt, aes(flux, prob))
p <- p + geom_area(aes(fill = flux > ci_flux))
p <- p + geom_line()
p
```

The figure above shows the specified distribution of flux values in relation to the calculated CI of 1.24 nmol/m^2^/s. This shows that 
`r round(dlnorm(1.24, meanlog = 0, sdlog = 1)*100, 0)`% of fluxes (shaded red) will be lower than the CI, which can be considered as a limit of detection for the chamber-analyser system. 


### Variation in CI with chamber height

Using this function, we can examine how CI varies with various properties of the chamber-analyser system. Firstly, we see how CI is influenced by chamber height.

```{r}
v_height <- set_units(seq(0.1, 0.9, by = 0.1), m)
l_dt <- lapply(v_height, function(i){get_dt_ci(height = i)})
dt_h <- data.table::rbindlist(l_dt)
p <- ggplot(dt_h, aes(flux, prob))
p <- p + geom_area(aes(fill = flux > ci_flux))
p <- p + geom_line()
p <- p + facet_wrap(~ height) + ggtitle("Chamber height in m")
p
```


## Uncertainty in Cumulative flux or Emission Factors

tbc